When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-07-05 code/Submissionv1.Rmd
commit changes

---
title: "When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in the Contemporary Press"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper seeks to explain how retired military officers have attempted to shape US public opinion through the publication of op-ed pieces in major newspapers. I construct and analyze a data set that captures more than 300 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1975, and ultimately find two trends that stand out. The first is that retired military officers are increasingly writing about topics that are outside of traditional security related topics. The second trend is that retired military officers are increasingly criticizing Presidential administrations, especially that of President Donald Trump, for reasons other than disputes regarding international uses of force. Given the polarizing political climate in the United States and that a Presidential election will occur this fall, these two trends are concerning from the perspective of civil-military relations and suggest that the norm of a professional, apolitical military establishment is under significant strain._

@@

On June 1 2020, the President of the United States, Donald Trump, after giving remarks at the Rose Garden of the White House in Washington D.C., walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and after several minutes, returned to the White House.

In the days and weeks that followed, a full blown civil-military relations ensued [@feaver_military_2020]. Several cabinet members and uniformed military leaders sought to distance themselves from the President's threat of invoking the Insurrection Act of 1807 to use active-duty forces to quell domestic riots that had initiated following the death of George Floyd in Minneapolis. The Secretary of Defense, Mark Esper, stated that he "did not support" use of the 1807 Insurrection Act to deploy active duty military to establish order, and General Mark Milley, the Chairman of the Joint Chiefs of Staff, told a graduating class from the National Defense University that he "should not have been [joining the President on the walk to St. John's Church]," and that his presence "created a perception of the military involved in domestic politics" [@esper_secretary_2020; @milley_official_2020]. Multiple retired American generals also spoke out against the Trump administration's response  But for those paying close attention to what these retired military officers said, it is clear that there was some noticeable variation in their remarks. While all of them expressed deep concern over the way in which the President was threatening to use the military [@brooks_dismay_2020; @dempsey_former_2020 ], some attacked the President in much more forceful and personal terms [@goldberg_james_2020; @mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through the use of opinion-editorial pieces. It is an interesting question worthy of examination for at least three reasons. First, even after leaving active duty service, military officers remain part of the military institution - one that maintains considerable trust and credibility among the American public. A 2018 Pew Research Center poll determined that 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. The second reason this question is important is because as an institution, the military prides itself on being nonpartisan and apolitical, an ideal that is undoubtedly under significant stress in an era of increased political polarization [@layman_party_2006; @hare_polarization_2014]. Third, this question is worth exploring because it may reveal the degree to which the first two factors - military credibility and political polarization - have influenced the concept and norm of professionalism among the military as an institution. All of these factors are worthy of exploration, especially ahead of what promises to be a very spirited presidential election this coming fall.

This paper proceeds in three parts. First, I briefly review the basis upon which the norm of apolitical and nonpartisan behavior of the military as an institution is based. Here I briefly establish a basis for political polarization in contemporary America as well as the high degree of credibility the military as an institution presently enjoys. Second, I describe the analytical approach and the data used in this project. Assembling an original data set that includes more than 300 opinion pieces published by retired military officers in major American newspapers since the late 1970s, I conduct a mixed-methods approach which ultimately reveals that although the vast majority of opinion pieces published in major newspapers seem to uphold the ideal of the military as apolitical and non-partisan, this trend has been weakening in recent years. The data shows that retired military officers are increasingly writing about topics that are not immediately related to national security, as well as increasingly criticizing if not altogether attacking political administrations more often than they did in previous eras. Third, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a highly polarized political climate. 
\singlespace
# The Non-Partisan Military Institution, Credibility, and Polarization  
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is deep. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington. In his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, Huntington asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced [@huntington_soldier_1957]. The second force was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society, and that these forces were at odds with each other. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given today's polarizing environment, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer's concern was that of military officers adopting an ethic of professionalism such that they would actually be more likely - not less likely - to become involved in political issues and concerns [@finer_man_1962, 22-23]. Finer was especially concerned that the adoption of professionalism alone would lead to military professionals believing that they knew what was best for their country, over and above that of the specific, elected civilian leaders in office. Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. Finer's suggestion was that because of the perils of professionalism as espoused by Huntington, what was really needed to keep the military out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the concept are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed in debates about civil-military relations. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific who is also well known for the controversy surrounding his conduct in the Korean War and his eventual dismissal by President Truman, told the cadets at West Point in May of 1962.

@@

Current military doctrine and practices likewise reinforce the notion of the military remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. General Martin Dempsey, in an opinion piece published in the Washington Post on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

@@

## Political Polarization, Public Opinion, and Military Expertise

Scholars of American Politics have studied political polarization in depth, and while they have offered differing explanations for the causes of polarization, ranging from race and religion [@abramowitz_great_2018], to debates about the size and role of government [@sides_two_2015], to fierce policy disagreements over policy [@mann_its_2012], virtually all agree that polarization, if not an entirely new phenomenon, has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014]. Hare and Poole are among scholars who argue that modern American political polarization began in the 1970s as a result of redistribution along racial lines following key pieces of legislation in the 1960s [@hare_polarization_2014]. They further argue that polarization disagreement has only intensified, something that a significant portion of modern Americans agree on [@schaeffer_far_nodate].

Within this polarized political environment, what role does the military play - particularly retired military officers who possess considerable military expertise? In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argued that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information. He is rather acknowledging that most citizens care most about the world that is not "beyond their personal experience" [@zaller_nature_1992]. His point is that elites play a significant role in helping to form and sway public opinion.

Michael Robinson further examined how elite cues are received and processed. Robinson argues that partisan groups, as a result of ongoing political polarization, are cognitively biased in distinct ways in how they receive and process information [@robinson_who_2019]. As a result, Robinson found that partisans will express confidence in the military in vastly differing ways even after neutral, non-partisan information about various types of military failure is presented to these groups [@robinson_who_2019]. If Republicans are less likely to express a reduction in confidence in the military as an institution than Democrats are, as Robinson alleges, then even accurate, unbiased information is potentially irrelevant, as these differing partisan groups are pre-programmed to selectively see a particular situation as they wish to see it [@robinson_who_2019]. His main point serves to warn us regarding confidence in the military as an institution: there may be real reasons to lose confidence in the military as a result of a particular incident, but partisan filters will prevent a reduction of confidence from actually being expressed.

Zaller's framework makes it clear that it is a perfectly reasonable assumption that retired military officers who choose to express their opinions through the form of opinion-editorial pieces constitutes a form of elite discourse. After all, retired military officers have at least twenty years of experience, and in the case of most multi-star generals or admirals, usually more than 30 and in some cases close to 40 years of military experience. Indeed, most retired senior generals and admirals have spent their professional careers solely in the military, and have extensive experience while serving in leadership positions of large organizations and, in many cases, during times of war. Retired military officers are thus elites in the field of national security, and in line with Zaller's framework, they possess the potential to significantly contribute to policy debates and thus, to influence public opinion. Yet we also know that the American public will not necessarily evaluate the opinions of these military experts objectively, but rather through pre-conceived and potentially partisan lenses [@robinson_who_2019].

@@

Given these three parameters -- that retired military officers are part of an institution that highly values professionalism and non-partisanship, that political polarization in America has increased since at least the 1970s, and that retired military officers are elites who possess the potential to sway public opinion -- I now aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened. To answer this question, I assemble and examine a data set that consists of opinion pieces authored by retired military officers in major US newspapers. My methodological approach involves two key steps. First, I determine if there have been shifts in the topics that retired military officers have been writing about. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed. In simple terms, I want to answer the questions, what are these retired military officers writing about, and through the lens of professionalism, should they be saying what they are saying?

@@

The dynamics of the military institution engaging the American public in a polarized environment remains ripe for future scholarship. In the author's opinion, whether the American military can, and to what degree it will, be able to remain professional as an institution remains to be seen. As political polarization becomes increasingly salient, several dynamics are unfolding that are worth noting, as their full impact has yet to materialize.

The first is that as political polarization worsens, the entire notion of loyalty is being reshaped, redefined, or at the very least, called into question. Consider that in June 2020, following the President's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and politically striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@line_letter_2020]. The letter's clear implication is that if you are serving in the Trump administration, and if you graduated from the US Military Academy, you have forsaken your oath of office, and by extension, the US Constitution, as well as the values that West Point stands for. Increasing political polarization promises to make these charges of misplaced loyalty more common. Scholarly literature has looked at how the military faces pressure stemming from divided institutional loyalties [@betts_soldiers_1991], but polarization promises to add even more pressure and scrutiny on the loyalty of individual officers and the military as a whole.

A second dynamic, that of maintaining credibility and trust with the American public, remains crucial for the military, but this dynamic will likely be tested in fierce ways should a polarized political climate endure. Even as high credibility with the American public can give the military support, it can also be abused, especially if military officers, to include retirees, use the high level of credibility they currently enjoy to take up and advocate for partisan positions, or worse, to jockey with civilian leaders for power [@brooks_shaping_2008-1]. Unfortunately, as this paper illustrates, there is evidence that mild aspects of this are occurring, but the extent to which it is, and its resulting impacts, remains an area for future research. Credibility is not something that should be shunned. The military as an institution, to include military retirees, holds relevant and military expertise that should be afforded to the American public at large. And yet, just as retired officers were skilled in the management of violence when they wore the uniform every day, retirement in a polarized era means that they must so also be managers of American trust. The promising news is, as this paper makes clear, the great proportion of retirees who are speaking out in various settings know and internalize this. Nonetheless, this will become more difficult to do in a polarized environment.

A third, and perhaps the most important dynamic of polarization as it concerns the military, is that the line between what is considered partisan or political and what is considered morally right and wrong is becoming seemingly blurred, if not altogether disappearing. Polarized sides tend to believe that each is acting "rightly" and that the other side is not just mistaken, but deeply wrong in a moral sense.

These three factors are difficult in a purely domestic political environment, and one does not have to imagine how the potential snares of a polarized political environment and military credibility stewarded poorly could impede the military's ability to do it's main task, fighting and winning this nation's wars. Indeed, the impact of these characteristics on our nation's military effectiveness must also be considered in further research. Regardless, the importance of individual moral courage will continue to grow in value as an attribute of military leaders. Its importance as a trait will likely continue to grow in the years ahead.

>>>>

---
title: "When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. Assembling and analyzing more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979 reveals that there has been an increasing trend in published opinion pieces which fail to uphold professional standards as defined by military norms. The paper further argues that this trend is primarily the result of two significant strains on the military's politically neutral culture: political polarization and military credibility. Finally, the paper develops a theory that shows how the forces of political polarization and military credibility interact, and argues that the norm of a professional, apolitical military will remain under further strain in the current US political environment._

@@

In late May and early June 2020, the American public was gripped by escalating racial tension and discussions about police brutality following the death of George Floyd, an unarmed black man, by police in Minneapolis. Protests, most of them peaceful, were accompanied in some cases by rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square.  Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

In the days and weeks that followed, a civil-military relations crisis ensued  [@feaver_military_2020]. The Secretary of Defense, Mark Esper, publicly disagreed with the President's threat to invoke the Insurrection Act of 1807 to use active-duty forces to quell domestic riots [@esper_secretary_2020], and General Mark Milley, the Chairman of the Joint Chiefs of Staff, told senior military officers graduating from the National Defense University that he "should not have been there" (accompanying the President on the walk to St. John's Church), as his presence "created a perception of the military involved in domestic politics" [@milley_official_2020].

Retired military officers were among the elites who weighed in to give their opinions not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America. But for those paying attention to what these retired officers said, it is also clear that there was substantial variation in the tone and tenor of each officer's remarks. While all of them expressed deep concern about the possibility of active-duty forces confronting peaceful protesters in clear violation of the First Amendment of the US Constitution, [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further, leveling direct criticisms of the President's fitness to be President [@goldberg_james_2020], his administration's broader policies on race [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through opinion-editorial publications. It is an interesting question worthy of examination because, as I argue, at least three important factors are interacting in unique ways to shape how and why retired military officers speak out on a variety of issues. These factors are professionalism, polarization, and credibility. In particular, polarization and credibility are creating tremendous strains on the military's professional nature, and in some cases, are causing words to become weaponized.

This paper proceeds in four parts. First, I briefly review the concepts of military professionalism, polarization, and credibility in order to orient the reader to the theory that I build in part two. In the second section, I build a theory in which I argue that the contemporary forces of polarization and military credibility can strain the concept of professionalism such that words and actions become weaponized to the point of being antithetical to the very concept of professionalism itself. Third, I turn to the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. While the data shows that the vast majority of these pieces are in fact in line with the notion of military professionalism, recent trends show that the retired military officers are weighing in on topics that fall outside of the realm of national security as well as expressing clear partisan preferences in their arguments. Fourth, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a setting where polarization is high and the institution commands wide public support.

\singlespace
# Military Professionalism
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is deep. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given the potential influences of polarization and credibility, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer's concern was that of military officers adopting an ethic of professionalism such that they would actually be more likely - not less likely - to become involved in political issues [@finer_man_1962, 22-23]. Finer was especially concerned that the adoption of professionalism alone would lead to long-serving military professionals believing that they knew what was best for their country over and above that of temporary elected civilian leaders [@finer_man_1962, 22-23] Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. Finer's suggestion was that because of the perils of professionalism as espoused by Huntington, what was really needed to keep the military out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the same are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed in debates about the perils of professionalism. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]  
It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions, albeit with a far less than perfect record. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific and who later commanded the occupation of Japan, is well known for his disagreement with President Truman regarding American strategic goals during the Koran War. Yet his remarks to cadets at West Point as an 82 year old retired officer are also well known and applicable to the topic at hand:

@@

Current military doctrine and practices likewise reinforce the notion of the military institution remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. Retired General Martin Dempsey, in an opinion piece published in _The Washington Post_ on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

@@

## Military Credibility and Expertise

Military credibility remains high among the US population, though this is not something that has come easy for the military. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Additionally, military credibility is high relative to other institutions. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].

When credibility between the American population and the military as in institution is high, it can be inferred that the public trusts the military - to accomplish its military mission most of all, but also to do more than that - to develop healthy institutional cultures, to vigorously address unethical conduct, and to exercise good leadership while taking care of American service members [@us_army_repairing_2014]. Senior military leaders are aware of the importance of maintaining public trust, and are generally sensitive to issues or concerns which may lead to a loss in confidence of the military [@us_army_repairing_2014].

Scholars have also shown that when elites possess a high degree of credibility, their ability to influence and shape public opinion is likewise high. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

## Political Polarization

Almost all accounts of polarization in the scholarly literature point to American society becoming increasingly polarized. While scholars have offered differing explanations for the causes of polarization, ranging from concerns over race and religion [@abramowitz_great_2018], to the size and role of government [@sides_two_2015], to fierce policy disagreements over policy issues [@mann_its_2012], virtually all scholars agree that polarization in America has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014], and by many measures, is worse than it has been during any period since the end of the American Civil War [@hare_polarization_2014]. Scholars have also shown that whereas polarization in previous decades was pronounced over singular and specific issue areas, to include redistribution policies following key Civil Rights legislation passed in the 1960s [@hare_polarization_2014], recent polarization is marked by a phenomenon known as "conflict extension," [@layman_party_2002], referring to the notion that a far wider set of issues and policy preferences covering the full gamut of social welfare, racial, and cultural questions are taking on a polarizing character [@layman_party_2006].

The 2016 Presidential Election, resulting in the Trump presidency, confirms the nature and extent of polarized politics in America. Political scientists, journalists, and reporters are still trying to piece together the specific factors that led to Donald Trump's surprise victory. Sides, Tesler, and Vavreck argue that Trump's election reveals just how split Americans are, particularly on core issues of identity and attitudes regarding race and religion [@sides_identity_2018]. During the years of the Trump Presidency, moreover, the American people have indicated that polarization is an issue whose presence is being felt and worsening [@schaeffer_far_2020]. Finally, given that Americans feel more strongly today about more issues and especially those issues from which individuals draw a sense of core identity, it is hardly surprising that individuals seem to apply partisan lenses, or biases, to what they hear. Michael Robinson showed as a result of political polarization, partisan groups are cognitively biased in how they receive and process information [@robinson_who_2019]. As a result, Robinson found that partisans will express confidence in the military in vastly differing ways even after neutral, non-partisan information about various types of military failure is presented to these groups [@robinson_who_2019].

## When Words Become Weaponized - Theory

The purpose of this section is to explain how the concepts of professionalism, credibility, and political polarization can interface to create a potentially dangerous atmosphere for both military professionals and the public. It must be stressed from the outset that none of these concepts in isolation impose a normative prohibition on the act of a military officer taking up the pen to write an opinion piece. Indeed, Huntington's concept of military professionalism recognizes the unique skills of military officers, and Zaller's framework of elite discourse tacitly affirms that retired military officers, especially generals and admirals, most of whom possess more than 30 and in some cases 40 years of professional experience, possess relevant expertise that the public would find helpful, relevant, and convincing as it pertains to a host of potential policy debates. Even the existence of political polarization itself is insufficient to suggest that retired military officers should never give their opinion on specific matters.

One can reasonably infer that a professional opinion piece written by a military officer would then take on three very important qualities. First, stemming from the concept of military professionalism, the piece would address a relevant military topic. To be clear, multistar generals and admirals often are experienced in the realm of strategy, a word that is often misunderstood and misapplied in discussions of the topic [see for example @strachan_direction_2014]. Without unpacking the concept of strategy further, the relevant point is that military elites, and especially those who achieved a high rank, have been privy to decisions and experiences occurring near the highest levels of government. Their body of experiences is not confined to mere tactics. They They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service. Second, from a credibility and expertise perspective, a professional piece would be based on the officer's personal experiences and would be written from an individual's perspective so as not to infer that an argument is being made on behalf of the collective service or element within the military body. Third, an opinion made by a military officer would be free of any hint of partisan sentiment, whether advocacy or denouncement. If one of these three qualities - addressing a relevant military topic, from the perspective of an individual, and free of partisan expression - fails to be upheld, there is increasing likelihood that the officer's words can serve as a weapon, and, like other kinds of weapons, can do great harm.

My theory holds that when military credibility is high, a demand signal is created by the American public such that future interactions between military elites and the military institution are encouraged and welcomed. There are numerous ways in which these interactions might satisfy the public's demand to observe or hear from the military. Institutionally, the military might respond to the public's demand signal for increased engagement by participating in public ceremonies, parades, and other showcases of military prowess (i.e. at sporting venues). Individually, military elites, especially retired military officers, would be more eager to dialogue with the American public when military credibility is high. Press and media outlets would be more eager to showcase elite military voices when military credibility is high because they represent an institution that commands a high degree of public respect. When military credibility is low, the opposite will occur. Since demand for interactions between the American public and the military is muted, the institution will not participate in a high number of public showcases of military prowess because there will not be as strong of a demand signal sent by the public. In short, military credibility and respect for military expertise impacts the likelihood of the military institution interacting with the American public. When respect for the military is high, the military is likely to speak out and to interact with the public, and when respect for the military is low, the military is less likely to do so.

When political polarization is high, I contend that it ultimately places great strain on the concept and definition of loyalty because the distinction between what is considered morally right or wrong and partisan expression becomes blurred if not altogether removed. The result is a less unified military institution. I further contend that these mechanisms work as the logical and expected outworking of the characteristics of the polarized American political environment. While scholarly literature has looked at how the military faces pressure stemming from divided institutional loyalties [@betts_soldiers_1991], the dynamic of polarization promises to add even more pressure and scrutiny on the loyalty of individual officers and the military as a whole.

I contend that the the line between what is considered by individuals to be morally right and wrong and sentiments of partisan expression becomes blurred in highly polarized environments. As discussed above, political polarization has caused Americans to move further apart ideologically on an increasing number of issues [@layman_party_2006], including on core attitudes regarding issues such as race and immigration that are central to one's identity [@sides_identity_2018]. In this environment, individuals believe the stakes are raised to a very high level for a number pivotal concerns. Individuals in a polarized environment, I contend, also believe there are enormous consequences if the specific side with which they ally on any given issue loses to the other side because the other side is not just simply mistaken, but deeply and morally wrong.

This dynamic impacts military officers in a unique way because military officers, unlike regular citizens, operate under strict codes of loyalty all of the time. The American military officer, for example, takes an oath to support and defend the Constitution of the United States. In a polarized environment, two different officers may adopt completely different stances on a contentious issue, each believing the other is morally wrong and operating outside of the framework of the constitution. Each officer, then, might accuse the other of failing to be loyal to the ideals of the military. As a result, when polarization is high, the likelihood of partisan expression is high because in doing so, members of the military institution may feel that doing so is the right and loyal thing to do.

As a quick case in point, consider that in June 2020, following President Trump's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and politically striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@line_letter_2020]. The letter's clear implication is that if you are serving in the Trump administration and if you graduated from the US Military Academy (i.e. Esper and Pompeo), you have forsaken your oath of office, the US Constitution, and the values that West Point stands for.

@@

I now aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened. To answer this question, I assemble and examine a data set that consists of opinion pieces authored by retired military officers in major US newspapers. My methodological approach involves two key steps. First, I determine if there have been shifts in the topics that retired military officers have been writing about. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed. In simple terms, I want to answer the questions, what are these retired military officers writing about, and through the lens of professionalism, should they be saying what they are saying?

@@

The dynamics of the military institution engaging the American public in a polarized environment remains ripe for future scholarship. In the author's opinion, whether the American military can, and to what degree it will, be able to remain professional as an institution remains to be seen. As political polarization becomes increasingly salient, several dynamics are unfolding, all of which deserve attention in future scholarship.

The first dynamic, that of maintaining credibility and trust with the American public, remains crucial for the military, but this dynamic will likely be tested in fierce ways should a polarized political climate endure. Even as high credibility with the American public can give the military support, it can also be abused, especially if military officers, to include retirees, use the high level of credibility they currently enjoy to take up and advocate for partisan positions, or worse, to jockey with civilian leaders for power [@brooks_shaping_2008-1]. Unfortunately, as this paper illustrates, there is evidence that mild aspects of this are occurring, but the extent to which it is, and its resulting impacts, remains an area for future research.

Credibility is not something that should be shunned. The military as an institution, to include military retirees, holds relevant and military expertise that should be afforded to the American public at large. And yet, just as retired officers were skilled in the management of violence when they wore the uniform every day, retirement in a polarized era means that they must also become managers of American trust. The promising news is, as this paper makes clear, the great proportion of retirees who are speaking out in various settings know and internalize this. Nonetheless, this will become more difficult to do in a polarized environment.

The second, and perhaps more important dynamic, is understanding the impacts of the polarized environment on the military's ability to perform its primary military task, i.e. fight and win the nation's wars. Scholars, military officers, and practitioners must better understand what impacts, if any, are placed on military and civilian leaders as a result of political polarization on the development and implementation of war goals, aims, and strategies, and how military leaders and civilian policymakers can overcome these hurdles to generate effective policies [@rapp_civil-military_2015].
